---
title: 'Lecture 11: Principal component and factor analyses'
output: html_document
date: "2025/4/30"
---

# Principal component analysis

## RMD_example 11.1: Read 分類熱門照片 (picture.csv)

```{r}
# import the csv file into R
#   note: stringsAsFactors=TRUE will screw up conversion to numeric!
picture<-read.csv("picture.csv", header=TRUE, fileEncoding="BIG5")
dim(picture)
head(picture)

# 原本"上傳時間"的單位為" UNIX timestamp x 10^-8"
# 將上傳時間做轉換，以"天數"為單位，並相較於最早上傳的時間，計算出每筆資料與最早上傳的時間差
picture[,c("上傳時間")]<-picture[,c("上傳時間")]*10^8
t0<-as.POSIXlt(as.numeric(min(picture[,c("上傳時間")])), origin="1970-01-01", tz="GMT")
tt<-as.POSIXlt(as.numeric(picture[,c("上傳時間")]), origin="1970-01-01", tz="GMT")
picture[,c("上傳時間")]<-as.numeric(difftime(tt, t0, unit="day"))

head(picture)
```

## RMD_example 11.2: PCA using the covariance matrix (original measurements)

```{r}
princov<-princomp(picture[,2:8], cor=F)
summary(princov)
print(princov$loadings, digits=7, cutoff=0.0000001)
#   show a scree plot.
screeplot(princov, type="line", main="Scree plot")
#   plot 1st vs. 2nd PCs
x<-princov$scores[,1]
y<-princov$scores[,2]
plot(x, y, xlab="PC 1", ylab="PC 2", main="PCA with Cov", type="n")
text(x, y, labels=picture[,1], cex=.7)
legend("topright", c("女生個人","男生個人"), pch ='12')
```

## RMD_example 11.3: PCA using the correlation matrix (standardized variables)

```{r}
princor<-princomp(picture[,2:8], cor=T)
summary(princor)
print(princor$loadings, digits=7, cutoff=0.0000001)
#   show a scree plot.
screeplot(princor, type="line", main="Scree plot")
#   plot 1st vs. 2nd PCs
x<-princor$scores[,1]
y<-princor$scores[,2]
plot(x, y, xlab="PC 1", ylab="PC 2", main="PCA with Cor", type="n")
text(x, y, labels=picture[,1], cex=.7)
legend("topright", c("女生個人","男生個人"), pch ='12')
```

## RMD_example 11.4: Factor analysis

We will perform a factor analysis with the number of factors = 3. 

```{r}
factfit <- factanal(picture[,2:8], factors=3, rotation="none")
print(factfit, digits=7, cutoff=0.0000001, sort=TRUE)
```

The factor loadings are rotated using the varimax method, and the factor scores are estimated using the regression method.  

```{r}
factfit <- factanal(picture[,2:8], factors=3, scores="regression", rotation="varimax")
print(factfit, digits=7, cutoff=0.0000001, sort=TRUE)

# Print out the factor scores:
#  Every observation (picture) has scores on all 3 factors!
head(factfit$scores, n=25)
```

To plot the scree plot, we can do

```{r}
# Install package "psy" first, then load them into R
library(psy)
scree.plot(factfit$correlation)
```

We then plot 1st vs. 2nd factor scores.

```{r}
x<-factfit$scores[,1]
y<-factfit$scores[,2]
plot(x, y, xlab="Factor score 1", ylab="Factor score 2", main="Factor analysis", type="n")
text(x, y, labels=picture[,1], cex=.7)
legend("topright", c("女生個人","男生個人"), pch ='12')
```

