---
title: 'Lecture 8: Linear regression analysis'
output: html_document
date: "2025/4/9"
---

# Linear regression

## RMD_example 08.1: Read 六都房地產實價登錄資料 (house.csv) 

```{r}
# Import the csv file into R
#   note: stringsAsFactors=TRUE will screw up conversion to numeric!
house <- read.csv("house.csv", header=TRUE, fileEncoding="BIG5")

dim(house)
head(house)
```

## RMD_example 08.2: Simple linear regression

Let's first examine the histogram of y (每平方公尺單價)

```{r}
# Histogram of 每平方公尺單價 with a smooth density estimate
hist(house$每平方公尺單價,
     main="Histogram for 每平方公尺單價",
     xlab="每平方公尺單價",
     ylab="",
     border="blue", 
     col="green",
     xlim=c(0, 300000),
     las=1,
     breaks=100,
     prob = TRUE)
lines(density(house$每平方公尺單價))
```

Apparently, the data is right skewed. A log transformation on y can make it symmetric (like normal).

```{r}
# Histogram of log(每平方公尺單價) with a smooth density estimate
hist(log(house$每平方公尺單價),
     main="Histogram for log(每平方公尺單價)",
     xlab="log(每平方公尺單價)",
     ylab="",
     border="blue", 
     col="green",
     xlim=c(7, 14),
     las=1,
     breaks=50,
     prob = TRUE)
lines(density(log(house$每平方公尺單價)))
```

We then perform the simple linear regression on y.

```{r}
# Summary statistics of 每平方公尺單價 and 屋齡
#   Install package "psych" first, then load it into R
library(psych)
每平方公尺單價 <- house[,"每平方公尺單價"]
屋齡 <- house[,"屋齡"]
describe(每平方公尺單價)
describe(屋齡)
cor.test(每平方公尺單價, 屋齡)

# Simple linear regression analysis of 每平方公尺單價 on 屋齡
lmJE <- lm(每平方公尺單價~屋齡)
summary(lmJE)
#   plots for regression diagnosis
par(mfrow=c(2,2))
plot(lmJE, main="每平方公尺單價 vs. 屋齡")

# Summary statistics of 每平方公尺單價 by 車位
車位 <- house[,"車位"]
車位[車位=="無"]<-0
車位[車位=="有"]<-1
車位 <- as.numeric(車位)
describeBy(每平方公尺單價, 車位)

# Simple linear regression analysis of 每平方公尺單價 on 車位
lmJm <- lm(每平方公尺單價~車位)
summary(lmJm)
#   plots for regression diagnosis
par(mfrow=c(2,2))
plot(lmJm, main="每平方公尺單價 vs. 車位")
```

Perform these simple linear regressions on log(y).

```{r}
# Simple linear regression analysis of log(每平方公尺單價) on 屋齡
lmlJE <- lm(log(每平方公尺單價)~屋齡)
summary(lmlJE)
#   plots for regression diagnosis
par(mfrow=c(2,2))
plot(lmlJE, main="log(每平方公尺單價) vs. 屋齡")

# Simple linear regression analysis of log(每平方公尺單價) on 車位
lmlJm <- lm(log(每平方公尺單價)~車位)
summary(lmlJm)
#   plots for regression diagnosis
par(mfrow=c(2,2))
plot(lmlJm, main="log(每平方公尺單價) vs. 車位")
```

## RMD_example 08.3: Multiple linear regression

```{r}
# Multiple linear regression analysis of 每平方公尺單價 on 車位 and 屋齡
lmJma<-lm(每平方公尺單價~車位+屋齡)
summary(lmJma)
#   plots for regression diagnosis
par(mfrow=c(2,2))
plot(lmJma, main="每平方公尺單價 vs. 車位+屋齡")
```

## RMD_example 08.4: Polynomial regression

```{r}
# Linear regression analysis of 每平方公尺單價 on 屋齡 and 屋齡^2
屋齡2<-屋齡*屋齡
lmJEE2<-lm(每平方公尺單價~屋齡+屋齡2)
summary(lmJEE2)
#   plots for regression diagnosis
par(mfrow=c(2,2))
plot(lmJEE2, main="每平方公尺單價 vs. 屋齡+屋齡2")
```

## RMD_example 08.5: Dummy variables

```{r}
# Linear regression analysis of 每平方公尺單價 on dummy variables of 區域
區域<-house[,"區域"]
X北<-區域
X北[!is.na(區域) & 區域=="台北市"]<-1
X北[!is.na(區域) & 區域!="台北市"]<-0
X北<-as.numeric(X北)
X新<-區域
X新[!is.na(區域) & 區域=="新北市"]<-1
X新[!is.na(區域) & 區域!="新北市"]<-0
X新<-as.numeric(X新)
X桃<-區域
X桃[!is.na(區域) & 區域=="桃園市"]<-1
X桃[!is.na(區域) & 區域!="桃園市"]<-0
X桃<-as.numeric(X桃)
X中<-區域
X中[!is.na(區域) & 區域=="台中市"]<-1
X中[!is.na(區域) & 區域!="台中市"]<-0
X中<-as.numeric(X中)
X南<-區域
X南[!is.na(區域) & 區域=="台南市"]<-1
X南[!is.na(區域) & 區域!="台南市"]<-0
X南<-as.numeric(X南)

lmJEd<-lm(每平方公尺單價~X北+X新+X桃+X中+X南)
summary(lmJEd)
#   plots for regression diagnosis
par(mfrow=c(2,2))
plot(lmJEd, main="每平方公尺單價 vs. dummy of 區域")
```

## RMD_example 08.6: Interaction in regression

```{r}
# Linear regression analysis for interactions
屋齡分組<-rep(0, length(屋齡))
屋齡分組[!is.na(屋齡) & 屋齡>25]<-1
屋齡分組[is.na(屋齡)]<-NA
車位X屋齡分組<-車位*屋齡分組

lmJmagmXag<-lm(每平方公尺單價~車位+屋齡分組+車位X屋齡分組)
summary(lmJmagmXag)
#   plots for regression diagnosis
par(mfrow=c(2,2))
plot(lmJmagmXag, main="interactions")
```

## RMD_example 08.7: Confounding in regression

```{r}
# Linear regression analysis of 每平方公尺單價 on 車位
lmJm <- lm(每平方公尺單價~車位)
summary(lmJm)

# Linear regression analysis of 每平方公尺單價 on 車位 and 屋齡分組
lmJmag<-lm(每平方公尺單價~車位+屋齡分組)
summary(lmJmag)
#   plots for regression diagnosis
par(mfrow=c(2,2))
plot(lmJmag, main="每平方公尺單價 vs. 車位+屋齡分組")
```

## RMD_example 08.8: Variable selection

```{r}
# All possible regressions
library(leaps)
車位X屋齡<-車位*屋齡
d<-cbind(每平方公尺單價, 車位, 屋齡, 屋齡2, 車位X屋齡)
for(i in 1:dim(d)[2]) d <- d[!is.na(d[,i]),]   # delete the NA's (missing values)
resall<-regsubsets(x=d[,(2:dim(d)[2])], y=d[,1],
                   nbest=1,         # 1 best model for each number of predictors
                   nvmax=NULL,      # NULL for no limit on number of variables
                   force.in=NULL, force.out=NULL,
                   method="exhaustive")
summary(resall)
#   plot results
library(car)
par(mfrow=c(1,2))
subsets(resall, statistic="adjr2", main="All: adjusted R^2", legend=c(1.5, 0.02))   # for adjusted R^2
subsets(resall, statistic="bic", main="All: BIC", legend=c(1.5, -100))              # for BIC

# Backward elimination
base<-lm(每平方公尺單價~1, data=data.frame(d))
full<-lm(每平方公尺單價~車位+屋齡+屋齡2+車位X屋齡, data=data.frame(d))
step(full, direction="backward", test="F")

# Forward selection
step(base, scope=list(lower=base, upper=full), direction="forward", test="F")

# Stepwise regression
step(base, scope=list(upper=full), direction="both", test="F")
```

