---
title: "Lecture 6: Nonparametric methods"
output: html_document
date: "2025/3/26"
---

# RMD_example 06.1: Read the Gene expression microarray data (samplexprs.csv) 
```{r}
# Import the csv file into R
#   note: stringsAsFactors=TRUE will screw up conversion to numeric!
samplexprs <- read.csv("samplexprs.csv", header=TRUE, stringsAsFactors=FALSE)

dim(samplexprs)
head(samplexprs[,1:10])

J00129<-samplexprs[,"J00129"]
metastases<-samplexprs[,"metastases"]
ERp<-samplexprs[,"ERp"]
Contig29982_RC<-samplexprs[,"Contig29982_RC"]
```

# One sample test

## RMD_example 06.2: One-sample t-test

```{r}
# Ointment data
Oint<-c(37,39,31,39,38,47,35,30,25,40)

# Summary statistics for Oint
#   Install package "psych" first, then load it into R
library(psych)
describe(Oint)

# One-sample t-test
t.test(Oint, mu=34, alternative="greater") 
```

## RMD_example 06.3: Sign test

```{r}
# Install the package "BSDA" first, then load it into R
library(BSDA)

SIGN.test(Oint, md=34, alternative="greater") 
```

## RMD_example 06.4: Sign test—for gene expression data 

```{r}
# Install the package "BSDA" first, then load it into R
library(BSDA)
SIGN.test(J00129, md=-0.5) # Ho: mu=-0.5
```

# Paired data

## RMD_example 06.5: Paired t-test

```{r}
# Twin paired data
twin1<-c(659,984,397,574,447,479,676,761,647,402)
twin2<-c(452,507,460,787,351,277,234,516,577,338)

# summary statistics for twin1 and twin2
describe(twin1)
describe(twin2)
describe(twin1-twin2)

# paired t-test for comparing twin1 vs twin2
t.test(twin1, twin2, paired=T, alternative="greater")
```

## RMD_example 06.6: Sign test for the difference twin1-twin2

```{r}
D<-twin1-twin2
SIGN.test(D, md=0, alternative="greater") # Ho: D_mu=0
```

## RMD_example 06.7: Wilcoxon signed-rank test

```{r}
# install the package "MASS" first, then load it into R
library(MASS)

# exact test
wilcox.test(twin1, twin2, paired=T, exact=T, alternative="greater")

# Normal approximation
wilcox.test(twin1, twin2, paired=T, exact=F, correct=F, alternative="greater")
```

## RMD_example 06.8: Wilcoxon signed-rank test—for gene expression data 

```{r}
wilcox.test(J00129, Contig29982_RC, paired=T)
```

# Two sample test 

## RMD_example 06.9: Two-sample t-test

```{r}
aspX<-c(15,26,13,28,17)
aspY<-c(12,20,10,21,18,22)

# summary statistics for aspX, aspY
describe(aspX)
describe(aspY)

# two-sample t-test for aspX vs. aspY
t.test(aspX, aspY, var.equal=T, alternative="greater")
t.test(aspX, aspY, var.equal=F, alternative="greater")
var.test(aspX, aspY)
```

## RMD_example 06.10: Wilcoxon rank-sum (Mann-Whitney) test

```{r}
# exact test
wilcox.test(aspX, aspY, paired=F, exact=T, alternative="greater")

# Normal approximation
wilcox.test(aspX, aspY, paired=F, exact=F, correct=F, alternative="greater")
```

## RMD_example 06.11: Wilcoxon rank-sum (Mann-Whitney) test—for gene expression data 

```{r}
wilcox.test(J00129~metastases)
```

# More than 2 samples

## RMD_example 06.12: ANOVA—for gene expression data 

```{r}
# summary statistics of J00129 (log10 gene expression intensity ratios) by ERp
describeBy(J00129, ERp)

# more than two samples test for J00129 over ER
anova(lm(J00129~as.factor(ERp)))
```

## RMD_example 06.13: Kruskal-Wallis test—for gene expression data 

```{r}
kruskal.test(J00129~as.factor(ERp))
```

# Correlation

## RMD_example 06.14: Pearson correlation—for gene expression data

```{r}
# correlation between two continuous gene expressions J00129 and Contig29982_RC
cor.test(J00129, Contig29982_RC, method="pearson")
```

## RMD_example 06.15: Spearman rank correlation—for gene expression data

```{r}
# correlation between continuous gene expression J00129 and ordinal expression level ERp
cor.test(J00129, ERp, method="spearman")
```
