---
title: 'Lecture 1-1: Introduction to R'
output: html_document
date: "2025/2/19"
---

# Getting started with R

## RMD_example 01-1.1: Installing the swirl package

```{r, eval=FALSE}
# install package from within R
install.packages("swirl")

# load the package into our R sessions for use
library(swirl)

# run the function
swirl()
```

## RMD_example 01-1.2: Use base functions and datasets

When you download R from CRAN you get what we call _base_ R. This includes several _functions_ that are considered fundamental for data analysis. It also includes several example datasets. These datasets are particularly useful as examples when we are learning to use the available functions. You can see all the available dataset by executing the function `data` like this:

```{r,eval=FALSE}
data()
```

Because in R functions are objects, we need the two parenthesis to let R know that we want the function to be executed as opposed to showing us the code for the function. Type the following and note the difference:

```{r,eval=FALSE}
data
```

To see na example of functions at work, we will use to `co2` dataset to illustrate the function `plot`, one of the base functions. We can plot Mauna Loa Atmospheric CO2 Concentration versus data like this:

```{r}
plot(co2)
```

Note that R's base functionality is bare bones. Note that data science applications are broad, the statistical toolbox is extensive, and most users need only a small fraction of all the available functionality. Therefore, a better approach is to make specific functionality available _on demand_.  R does this using _packages_, also called _libraries_. 

Some packages are considered popular enough that they are included with the base download.
For example, the software implementing the method of survival analysis are in the `survival` package. To bring that functionality to your current session we type

```{r,eval=FALSE}
library(survival)
```

## RMD_example 01-1.3: Installing packages

Installing R only includes a basic set of functions.
However, there are over 4,000 packages that are not included in the base installation. 
Many of these functions are stored in CRAN where
packages are vetted: they are checked for common errors and they must have a dedicated maintainer. 
You can easily install CRAN packages from within R if you know the name of the packages. As an example, we are going to install the package `dplyr` which we use in our first data analysis examples: 

```{r,eval=FALSE}
install.packages("dplyr")
```

We can then load the package into our R sessions using the `library` function:

```{r, ,eval=FALSE}
library(dplyr)
```

From now on you will see that we sometimes load packages without
installing them. This is because once you install the package, it
remains in place and only needs to be loaded with `library`. If you
try to load a package and get an error, it probably means you need to
install it first.

## RMD_example 01-1.4: Mathematics in R Markdown

In side a text chunk of R Markdown, you can use mathematical notation if you surround it by dollar signs `$` for inline mathematics and `$$` for displayed equations. 

Example: `$\sum_{n=1}^{10} n^2$` is rendered as $\sum_{n=1}^{10} n^2$

Example: `$$\sum_{n=1}^{10} n^2$$` is rendered as $$\sum_{n=1}^{10} n^2$$

The mathematical typesetting is based on $\LaTeX$, so if you need to search for the way to make a particular symbol, include latex in your search. But note: Not all $\LaTeX$ macros are available without using additional packages. 