---
title: "Lecture 5: Measures of association and hypothesis testing for categorical data"
output: html_document
date: "2025/3/19"
---

# RMD_example 5.1: Read the Gene expression microarray data (samplexprs.csv) 

```{r}
# Import the csv file into R
#   note: stringsAsFactors=TRUE will screw up conversion to numeric!
samplexprs <- read.csv("samplexprs.csv", header=TRUE, stringsAsFactors=FALSE)

dim(samplexprs)
head(samplexprs[,1:10])

J00129<-samplexprs[,"J00129"]
metastases<-samplexprs[,"metastases"]
ERp<-samplexprs[,"ERp"]
Contig29982_RC<-samplexprs[,"Contig29982_RC"]
```

# RMD_example 5.2: Binomial test

```{r}
# generating frequency tables
#   install package "gmodels", "epiR" first, then load them into R
library(gmodels)
library(epiR)

#   create ERs=1 if ERp>10, 0 o.w.
ERs<-rep(0, length(ERp))
ERs[!is.na(ERp) & ERp>10]<-1
ERs[is.na(ERp)]<-NA

# test for a single proportion
TAB<-xtabs(~ERs)
TAB
#   notice: p=Pr(ERs=0)
chisq.test(TAB) # Ho: p=0.5
chisq.test(TAB, p=c(0.4, 0.6)) # Ho: p=0.4
binom.test(TAB, p=0.5) # Ho: p=0.5
binom.test(TAB, p=0.4) # Ho: p=0.4
```

# RMD_example 5.3: Cohort and case-control studies

For exploring the data, we first create the $2 \times 2$ table.

```{r}
TAB1<-xtabs(~ERs+metastases)
TAB1
```

If the data are from a chort study, we can have relative risk and odds ratio. In the function `epi.2by2`, notice that "Exposed +" is for "ERs=0" and "Exposed -" is for "ERs=1", and "Outcome +" is for "metastases=0" and "Outcome -" is for "metastases=1"

```{r}
epi.2by2(TAB1, method="cohort.count")
```

If the data are from a case-control study, we can have the odds ratio.

```{r}
epi.2by2(TAB1, method="case.control")
```

# RMD_example 5.4: $\chi^{2}$ / Fisher's exact test

```{r}
# chi2 / Fisher's exact test
CrossTable(ERs, metastases, prop.chisq=F, chisq=T, fisher=T)
```

# RMD_example 5.5: McNemar's test for matched-pair studies

```{r}
# poor prognosis matches good prognosis 
Gg<-c(1,0,1,1,1,1,1,0,0,1,1,0,1,1,1,1,1,1,1,1,1,1,0,1,0)
Pg<-c(0,1,0,1,1,1,0,0,0,0,1,1,0,1,1,1,1,1,1,1,1,1,1,0,0)
TAB2<-xtabs(~Gg+Pg)
TAB2

# McNemar's test for ERs on prognosis
mcnemar.test(TAB2)
```

# RMD_example 5.6: Measure of agreement and its test

```{r}
# measure of agreement
#    on expression of J00129 and Contig29982_RC
Jexpry<-rep(0, length(J00129))
Jexpry[!is.na(J00129) & (J00129 > -0.5)]<-1
Jexpry[is.na(J00129)]<-NA
Cexpry<-rep(0, length(Contig29982_RC))
Cexpry[!is.na(Contig29982_RC) & (Contig29982_RC > -0.5)]<-1
Cexpry[is.na(Contig29982_RC)]<-NA
TAB3<-xtabs(~Jexpry+Cexpry)
TAB3

# kappa coefficient/test
#   install the package "fmsb" first, then load it into R
library(fmsb)
Kappa.test(TAB3)
```

# RMD_example 5.7: Interaction

```{r}
# create age45=1 if age>45, 0 o.w.
age<-samplexprs[,"age"]
age45<-rep(0, length(age))
age45[!is.na(age) & age>45]<-1
age45[is.na(age)]<-NA

TAB4<-xtabs(~ERs+metastases+age45)
TAB4
```

# RMD_example 5.8: Mantel-Haenszel test

```{r}
# smoking versus lung cancer for different genders
TAB5<-array(c(36,131,4,29,
              140,27,60,23),
            dim=c(2,2,2))
dimnames(TAB5)<-list(c("Non-smokers","Smokers"), c("Lung cancer no","Lung cancer yes"), c("Males","Females"))
TAB5
epi.2by2(TAB5, method="cohort.count")
mantelhaen.test(TAB5)
```

